---
title: "Mouse Liver Digestion Buffer"
subtitle: "Abdullah Lab, IMMEI, University Hospital Bonn"
protocol_id: "BUF-LIV-DIG"
version: "v1.0"
date: "2026-08-05"
description: "Collagenase and DNase I digestion buffer for enzymatic dissociation of mouse liver tissue. Supports Collagenase D or Collagenase IV in a DMEM, PBS, or HBSS base."
---

**Protocol ID:** {{< meta protocol_id >}}
**Version:** {{< meta version >}}
**Author:** Dillon Corvino

# Purpose

This buffer is used for enzymatic digestion of mouse liver tissue to generate viable single-cell suspensions. It contains a collagenase and DNase I at working concentrations suitable for dissociating liver parenchyma while preserving lymphocyte viability.

This single buffer replaces the former separate Collagenase D and Collagenase IV documents (`BUF-LIV-DIG-COL-D`, `BUF-LIV-DIG-COL-IV`, both retired at v2.0 of `MUS-LIV-DIG`).

\tableofcontents

# Linked protocols

```{r}
#| echo: false
#| message: false

library(tidyverse)

this_buffer <- rmarkdown::metadata$protocol_id
deps <- read_csv(here::here("protocols_src/_dependencies.csv"), show_col_types = FALSE)

linked <- deps |>
  filter(buffer_id == this_buffer) |>
  distinct(protocol_id)

if (nrow(linked) == 0) {
  cat("None\n")
} else {
  linked |>
    mutate(line = glue::glue("- `{protocol_id}`")) |>
    pull(line) |>
    cat(sep = "\n")
}
```

---

# Choice of collagenase and base medium

Two collagenases are in routine use. They are **not interchangeable at the same concentration** — Collagenase D is used at **twice** the concentration of Collagenase IV.

| Enzyme | Working concentration | Notes |
|---|---|---|
| **Collagenase D** | **1 mg/mL** | Roche 11088858001; supplied as lyophilised powder |
| **Collagenase IV** | **0.5 mg/mL** | Half the concentration used for Collagenase D |

**DNase I is 50 ug/mL for both.**

The base medium may be **DMEM (incomplete)**, **PBS 1x**, or **HBSS 1x** — these are interchangeable for this application. Use whichever is on hand; HBSS and PBS avoid the phenol red background in DMEM if that matters downstream.

::: callout-important
Do **not** use an EDTA-containing buffer (e.g. MACS Buffer) as the base medium, and do not perfuse the liver with one. Collagenases are Ca2+/Zn2+-dependent metalloproteases and EDTA chelates these cofactors, inhibiting digestion.
:::

# Working buffer composition

```{r}
#| echo: false
#| message: false

library(tidyverse)

# ---- inputs (edit here) ----
collagenaseStockMgMl <- 100      # mg/mL
dnaseStockMgMl       <- 10       # mg/mL
dnaseFinalUgMl       <- 50       # ug/mL, both enzymes
volumesMl            <- c(1, 5, 10, 25, 50)
enzymes.df <- tibble::tibble(
  enzyme      = c("Collagenase D", "Collagenase IV"),
  finalMgMl   = c(1, 0.5)
)
# ----------------------------

recipe.df <- tidyr::expand_grid(enzymes.df, volumeMl = volumesMl) |>
  dplyr::mutate(
    collagenaseUl = finalMgMl * volumeMl / collagenaseStockMgMl * 1000,
    dnaseUl       = (dnaseFinalUgMl / 1000) * volumeMl / dnaseStockMgMl * 1000
  )

# verify the arithmetic reproduces the target concentrations exactly
stopifnot(all(abs(recipe.df$collagenaseUl / 1000 * collagenaseStockMgMl /
                    recipe.df$volumeMl - recipe.df$finalMgMl) < 1e-9))
stopifnot(all(abs(recipe.df$dnaseUl / 1000 * dnaseStockMgMl /
                    recipe.df$volumeMl * 1000 - dnaseFinalUgMl) < 1e-9))

recipe.df |>
  dplyr::mutate(
    baseMedium    = paste0("to ", scales::comma(volumeMl), " mL"),
    collagenaseUl = paste0(scales::comma(collagenaseUl), " uL"),
    dnaseUl       = paste0(scales::comma(dnaseUl), " uL")
  ) |>
  dplyr::select(enzyme, volumeMl, baseMedium, collagenaseUl, dnaseUl) |>
  knitr::kable(
    align = "l",
    col.names = c("Enzyme", "Final volume (mL)", "Base medium",
                  "Collagenase (100 mg/mL stock)", "DNase I (10 mg/mL stock)")
  )
```

**5 mL of working buffer is required per liver.**

# Preparation

## Stock solutions

| Reagent | Stock concentration | Storage |
|---|---|---|
| Collagenase D or IV | 100 mg/mL in sterile PBS | Aliquot at -20 °C |
| DNase I | 10 mg/mL | Aliquot at -20 °C; avoid repeated freeze-thaw |

## Reconstitution of Collagenase powder

Collagenase D is supplied as a **lyophilised powder** and must be reconstituted before use.

1. Add **1 mL sterile PBS 1x** directly to a vial containing **100 mg** Collagenase powder.
2. Swirl gently until fully dissolved; **do not vortex**.
3. This yields a **100 mg/mL** stock.
4. Aliquot into single-use volumes and store at **-20 °C**, labelled with enzyme, concentration, date, and initials.

## Preparation of working buffer

1. Thaw collagenase and DNase I aliquots on ice.
2. Add the base medium (DMEM, PBS, or HBSS) to a sterile tube.
3. Add collagenase and DNase I per the table above.
4. Mix gently by inversion; **do not vortex**.
5. Pre-warm to **37 °C** immediately before use.

::: callout-note
Collagenase-only buffer (without DNase) may be prepared in advance and frozen in aliquots. Add DNase I fresh on the day of use.
:::

# Storage and stability

- Keep freshly prepared working buffer **on ice** until pre-warming for use.
- Do **not** freeze buffer after DNase I has been added.
- Discard unused working buffer at the end of the day.

# Materials

| Reagent | Supplier | Cat. # |
|---|---|---|
| Collagenase D | Roche, Basel, Switzerland | 11088858001 |
| Collagenase IV | [TBD] | [TBD] |
| DNase I | Sigma-Aldrich, Merck | DN25-100MG |
| DMEM (incomplete) | [TBD] | [TBD] |
| HBSS 1x | PAN-Biotech, Aidenbach, Germany | P04-34500 |
| DPBS, no Ca / no Mg | Gibco, ThermoFisher | 14190-094 |

# Version history

| Version | Date | Author | Change summary |
|---|---|---|---|
| v1.0 | 2026-08-05 | Dillon Corvino | Created by merging `BUF-LIV-DIG-COL-D` and `BUF-LIV-DIG-COL-IV`. DNase I raised to 50 ug/mL for both enzymes (was 10 ug/mL for Col D, 25 ug/mL for Col IV). Base medium stated as DMEM, PBS, or HBSS. |
