---
title: "37% Percoll Solution (= 33% final Percoll)"
subtitle: "Abdullah Lab, IMMEI, University Hospital Bonn"
protocol_id: "BUF-PERC-37"
version: "v2.0"
date: "2026-08-05"
description: "Percoll density solution for enrichment of leukocytes from digested liver tissue. 37% of isotonic Percoll stock, equivalent to 33% final Percoll."
---

**Protocol ID:** {{< meta protocol_id >}}
**Version:** {{< meta version >}}
**Author:** Dillon Corvino

# Purpose

This buffer is used for density-based enrichment of leukocytes from digested liver tissue. During centrifugation, hepatocytes and debris remain in the supernatant, while leukocyte-enriched pellets form at the bottom of the tube.

# "37%" and "33%" are the same solution

This buffer is called **37%** because it is **37% of isotonic Percoll stock**. The identical solution is **33% Percoll** when quoted by final Percoll content. Both numbers are correct — they use different denominators.

**SIP** — *Stock Isotonic Percoll*, also written *Standard Isotonic Percoll* — is the conventional term for neat Percoll pre-adjusted to physiological osmolality with 10x PBS at **9:1**. That mix is **90% Percoll by volume**, so:

$$\text{final Percoll \%} = \text{SIP \%} \times 0.9$$

```{r}
#| echo: false
#| message: false

library(tidyverse)

# ---- inputs (edit here) ----
percollInSip <- 0.9    # SIP = 9 parts neat Percoll : 1 part 10x PBS
sipPct       <- 37     # this buffer, expressed as % of isotonic stock
# ----------------------------

naming.df <- tibble::tibble(
  convention  = c("% of isotonic stock (SIP)", "% final Percoll"),
  thisBuffer  = c(sipPct, sipPct * percollInSip),
  denominator = c("Volume of SIP as a fraction of the working solution",
                  "Volume of neat Percoll as a fraction of the working solution")
)

stopifnot(abs(naming.df$thisBuffer[2] - 33.3) < 1e-9)

naming.df |>
  dplyr::mutate(thisBuffer = scales::percent(thisBuffer / 100, accuracy = 0.1)) |>
  knitr::kable(align = "l",
               col.names = c("Convention", "This buffer", "What the % is measured against"))
```

::: callout-important
When comparing against an external protocol, **check which convention it uses before changing any number**.

-   This buffer and `MUS-LIV-DIG` quote **% of isotonic stock**.
-   The AG Surace SOP #001 quotes **% final Percoll** — its "33% Percoll" (9 mL Percoll + 1 mL 10x DPBS + 17 mL 1x DPBS) is **the same density as this buffer**.

Relabelling "37%" as "33%" without also changing the recipe would drop the gradient to **29.7% final Percoll** and increase hepatocyte carryover.
:::

### Density vs osmolality (important conceptual note)

Percoll provides **density**, not physiological salt content.  
On its own, Percoll is **hypotonic** and unsuitable for live cells.

- **Density** determines where cells separate during centrifugation.
- **Osmolality** determines whether cells survive the procedure.

To maintain **physiological (1× PBS) osmolality**, Percoll must be rendered **isotonic** by adding concentrated salts (10× PBS).  
Once Percoll stock has been adjusted to isotonic conditions, **all subsequent working solutions can be prepared by dilution with 1× PBS only**.

\tableofcontents

# Linked protocols

```{r}
#| echo: false
#| message: false

library(tidyverse)

this_buffer <- rmarkdown::metadata$protocol_id
deps <- read_csv(here::here("protocols_src/_dependencies.csv"), show_col_types = FALSE)

linked <- deps |>
  filter(buffer_id == this_buffer) |>
  distinct(protocol_id)

if (nrow(linked) == 0) {
  cat("None\n")
} else {
  linked |>
    mutate(line = glue::glue("- `{protocol_id}`")) |>
    pull(line) |>
    cat(sep = "\n")
}
```

---

# Preparation of isotonic Percoll stock (required once per bottle)

Percoll stock **must be rendered isotonic before use**. This step is performed **once per bottle** and must be done under **sterile conditions**.

## Preparation from a 1000 mL Percoll bottle

1. Under a **biosafety cabinet**, using sterile technique:
   - Remove **100 mL Percoll** from the original 1000 mL bottle.
   - Transfer this 100 mL into a **clean, sterile glass bottle** (retain for later use).
2. To the original Percoll bottle, add **100 mL sterile 10× PBS**.
3. Cap securely and mix thoroughly by gentle inversion.
4. Clearly label the bottle with:
   - **“Isotonic Percoll”**
   - **“+100 mL 10× PBS”**
   - Date of preparation
   - Name or initials
   - **“Sterile”**

The remaining volume (~1000 mL) is now **isotonic Percoll stock**.

## Making the reserved 100 mL Percoll aliquot isotonic

1. To the previously removed **100 mL Percoll aliquot**, add:
   - **11.11 mL sterile 10× PBS**
2. Mix gently by inversion until homogeneous.
3. Label clearly with:
   - **“Isotonic Percoll”**
   - **“+11.11 mL 10× PBS”**
   - Date
   - Name or initials
   - **“Sterile”**

This aliquot is now **functionally identical** to the main isotonic Percoll stock and can be used interchangeably.

:::callout-warning
Percoll stock and all derived solutions **must be kept sterile**.  
:::

---

# Important note on isotonic Percoll usage

All working Percoll solutions in this protocol assume that **input Percoll is already isotonic**.

Therefore:
- **Do not add 10× PBS** when preparing 37% Percoll working solutions.
- Only **dilution with sterile 1× PBS** is required.

---

# Composition

| Component | Stock | Volume for 10 mL | Final composition | Notes |
|------------------|-------|------------------|-------------------|-------|
| Percoll (isotonic) | 100% | **3.70 mL** | 37% of SIP = **33.3% final Percoll** | Pre-adjusted stock |
| PBS 1× | — | **6.30 mL** | — | Diluent |

# Preparation of working solution

1. Equilibrate **isotonic Percoll stock** to **room temperature**.
2. Combine isotonic Percoll and sterile 1x PBS in the ratio **3.70 : 6.30**, scaled per the table below.
3. Mix gently by inversion until homogeneous.
4. Aliquot immediately into labelled tubes — for `MUS-LIV-DIG`, **25 mL per 50 mL conical tube**, one per liver.

Volumes below are the exact requirement per number of livers — see the tip beneath the table on adding margin.

```{r}
#| echo: false
#| message: false

library(tidyverse)

# ---- inputs (edit here) ----
percollInSip   <- 0.9    # SIP is 90% neat Percoll by volume
sipFraction    <- 0.37   # 37% of isotonic stock
volPerLiver    <- 25     # mL loaded per sample (MUS-LIV-DIG STEP 6)
nLivers        <- c(1, 2, 3, 4, 5, 6, 8, 9, 10, 12)
# ----------------------------

percoll.df <- tibble::tibble(nLivers = nLivers) |>
  dplyr::mutate(
    totalMl = nLivers * volPerLiver,
    sipMl   = totalMl * sipFraction,
    pbsMl   = totalMl - sipMl
  )

# components sum to total, and both percentages hold exactly at every scale
stopifnot(all(abs(percoll.df$sipMl + percoll.df$pbsMl - percoll.df$totalMl) < 1e-9))
stopifnot(all(abs(percoll.df$sipMl / percoll.df$totalMl - sipFraction) < 1e-9))
stopifnot(all(abs(percoll.df$sipMl * percollInSip / percoll.df$totalMl -
                    sipFraction * percollInSip) < 1e-9))

percoll.df |>
  dplyr::mutate(
    finalPct = scales::percent(sipMl * percollInSip / totalMl, accuracy = 0.1),
    dplyr::across(c(totalMl, sipMl, pbsMl), ~ round(.x, 2))
  ) |>
  knitr::kable(
    align = "l",
    col.names = c("Livers", "Total (mL)", "Isotonic Percoll (mL)",
                  "PBS 1x (mL)", "Final Percoll")
  )
```

Per liver this is **9.25 mL isotonic Percoll + 15.75 mL PBS** = 25 mL.

::: callout-tip
The table gives the **exact** volume for the number of livers being processed, with no margin built in. When preparing for more than one liver, **add some extra** so the last tube is not short — tube dead volume, pipetting loss, and the meniscus left in the stripette all add up across samples.

How much extra is your call: one additional liver's worth (25 mL) is usually sufficient, and is simplest to work out. For example, for 9 livers read off the 10-liver row.
:::

::: callout-note
That 10 : 17 ratio is arithmetically **37.04% SIP / 33.33% final** rather than exactly 37.00% / 33.30% — a difference of 0.03 percentage points, which is not meaningful at the bench. It is also **exactly** the AG Surace SOP recipe (9 mL Percoll + 1 mL 10x DPBS + 17 mL 1x DPBS), confirming the two protocols use the same gradient.
:::

::: callout-note
Prepare the working solution fresh on the day of use for optimal separation performance and cell viability.
:::

---

# Storage and stability

- **37% Percoll working solution**: prepare fresh on the day of use.
- Same-day storage at **room temperature** is acceptable.
- Do not freeze working solutions.
- **Isotonic Percoll stock** should be stored according to manufacturer’s instructions and maintained under sterile conditions.

---

# How isotonic Percoll calculations work (reference)

This section explains why **10× PBS** is added to Percoll and how the required volume is calculated.

## Definitions

- **V_T** = total final volume  
- **V_P** = volume of Percoll (contains no salts)  
- **V_10x** = volume of 10× PBS  

The remaining volume is **1× PBS**.

## Salt balance logic

The final solution must contain the same total amount of salt as **V_T mL of 1× PBS**.

Salt contributions:
- Percoll contributes **0×**
- 1× PBS contributes **1×**
- 10× PBS contributes **10×**

### Step 1: Express total salt content

```math
(V_T - V_P - V_10x) + 10 · V_10x = V_T
```

### Step 2: Combine terms

```math
V_T - V_P + 9 · V_10x = V_T
```

### Step 3: Subtract V_T from both sides

```math
- V_P + 9 · V_10x = 0
```

### Step 4: Rearrange

```math
9 · V_10x = V_P
```

### Final rule

```math
V_10x = V_P / 9
```

**In words:**  
For every **9 mL of Percoll**, add **1 mL of 10× PBS** to make the solution isotonic.

---

# Reagent details

| Component | Supplier | Cat# | Notes |
|----------------|----------|------|-------|
| Percoll (100%) | TBD | TBD | Must be rendered isotonic and kept sterile |
| PBS 1× | TBD | TBD | Sterile; Ca²⁺/Mg²⁺-free recommended |
| PBS 10× | TBD | TBD | Used once for isotonic adjustment |

---

# Safety (brief)

- Handle Percoll and PBS solutions with standard laboratory PPE.
- Dispose of Percoll-containing waste according to chemical and biological safety regulations.
- Avoid skin and eye contact; consult manufacturer SDS for details.

---

# Version history

| Version | Date | Author | Change summary |
|---------|------------|----------------|------------------------------------------------|
| v1.0 | 2025-11-20 | Dillon Corvino | Initial buffer definition (33% Percoll) |
| v1.1 | 2025-12-29 | Dillon Corvino | Updated to 37% Percoll; clarified isotonic prep |
| v1.2 | 2025-12-29 | Dillon Corvino | Added stock isotonic prep, math explanation, sterility notes |
| v2.0 | 2026-08-05 | Dillon Corvino | Density unchanged. Added explicit dual-convention section: 37% of isotonic stock = 33% final Percoll, with the AG Surace SOP #001 equivalence. Volumes for `MUS-LIV-DIG` raised to 25 mL per 50 mL tube (was 10 mL per 15 mL tube); scaling table added. |
